Multi-level modelling with Openflexo/FML - A contribution to the \mpc

### 2021-05-27 6-ModelPresentation.tex

Figure \ref{fig:MultilevelArchitecture} shows the overview of the multivel architecture designed to capture the two use cases described in the Process Challenge.

@@

Figure \ref{fig:BaseMetamodel} represents this base metamodel with a UML-like formalism, well adapted to represent object-oriented FML concepts and their instances. Base conceptual element is called a \textit{FlexoConcept} in FML language, and is similar to the concept of class in object-oriented modeling. We use the UML representation of class to represent a \textit{FlexoConcept}, while cardinalities are shown using the UML-like classical rules.

@@

\todo{Augmenter la taille des fontes de la figure \ref{fig:AcmeFullArchitecture}}

@@

Ici: un paragraphe qui décrit l'exécution, l'activation et la finalisation des taches, la business logic des gateways, etc...

@@

Figure \ref{fig:ToolingArchitecture} shows the architecture of these two tools. Left of figure presents \textit{process type graphical editor}. \textit{ProcessTypeEditor} is implemented as a \textit{VirtualModel} declaring two model slots (represented with bold circles). The first model slot references an instance of Acme metamodel, while the second model slot references an instance of diagram, conforms to the \textit{ProcessType diagram specification} (a diagram "metamodel" which defines and specify structure and graphical representations for edited items). While executed, this tool manages a graphical view for an instance of Acme metamodel (the \textit{Acme model}) and a specific \textit{ProcessType} instance. This tools allows to represent and edit a \textit{ProcessType}. One particular point should be noted regarding the highly reflective nature of FML and associated tooling : when drag and drop interactor is applied for a new item, the tool allows to choose the concept type to be instantiated (when a \textit{TaskType} is created for example, the user must choose the sub-concept of \textit{TaskType} to be selected - in Acme use case it can be \textit{SETaskTask} or \textit{CodingTaskType} or default value \textit{TaskType}).

The other tool, an \textit{enacted process editor}, provides process edition and tasks assignations through a graphical visualization displaying process being executed. This tool is represented one the right side of the figure \ref{fig:ToolingArchitecture}. It follows the same architecture pattern as explicited for the \textit{process type graphical editor}, with two model slots referencing both the model and a diagram. Process enactment is operated from a \textit{ProcessType}, and must be defined using a name identifying the newly instantiated process. All tasks are created from their \textit{TaskType} definition, and required assignments applies (if for example "Coding" \textit{TaskType} defines COBOL as output programming language, related task is automatically assigned to "Ann Smith"). All tasks get a status (\textit{Not startable} - when not assigned or when required input artifacts are not available - , \textit{Startable}, \textit{Started}, \textit{Completable}, \textit{Completed}), as well as a set of performing actors, a begin and end date, and used and produced artifacts. This tools also manages creation of artifacts, whose semantics follows rules defined in section \ref{sec:ProcessEnactment}}

% Expliquer l'outillage construit et comment il marche

>>>>

The designed multilevel architecture is shown in Figure \ref{fig:MultilevelArchitecture} overview and captures the two use cases described in the Process Challenge.

@@

Figure \ref{fig:BaseMetamodel} represents this base metamodel with a UML-like formalism, well adapted to represent object-oriented FML concepts and their instances. Base conceptual element is called a \textit{FlexoConcept} in FML language, and is similar to the concept of class in object-oriented modeling. The UML representation of class is used to represent a \textit{FlexoConcept}, while cardinalities are shown using the UML-like classical rules. 
In the diagram, we choose to represent the role associated with the \textit{FlexoConcept} in the class and the role type is referenced by the arrow in front of the role name.

@@

\todo[inline]{Augmenter la taille des fontes de la figure \ref{fig:AcmeFullArchitecture}}
\todo[inline]{Ajouter les programming languages}

@@

Base metamodel has been completed with some behaviours implementing execution semantic for executed processes. All tasks - instantiated from \textit{TaskType} for a given enacted process - manage a status which can be \textit{Not startable} (when not assigned to a performing actor or when required input artifacts are not available) , \textit{Startable}, \textit{Started}, \textit{Completable} (when all output artifacts are ready), \textit{Completed}). A \textit{Task} also manages a set of performing actors, a begin and end date, some used and produced artifacts. \textit{Gateways} business logic has also been implemented through the implementation of abstract \texttt{execute(Process)} behaviour. Management of artifacts whose semantics follows rules defined in section \ref{sec:ProcessEnactment} (\textbf{P13}) has also been implemented.

@@

Figure \ref{fig:ToolingArchitecture} shows the architecture of these two tools. Left of figure presents \textit{ProcessType graphical editor}. \textit{ProcessTypeEditor} is modeled as a \textit{VirtualModel} declaring two model slots (represented with bold circles). The first model slot references an instance of Acme metamodel, while the second model slot references an instance of diagram, conforms to the \textit{ProcessType diagram specification} (a diagram "metamodel" which defines and specify structure and graphical representations for edited items). While executed, this tool manages a graphical view for an instance of Acme metamodel (the \textit{Acme model}) and a specific \textit{ProcessType} instance. This tools allows to represent and edit a \textit{ProcessType}. One particular point should be noted regarding the highly reflective nature of FML and associated tooling : when drag and drop interactor is applied for a new item, the tool allows to choose the concept type to be instantiated (when a \textit{TaskType} is created for example, the user must choose the sub-concept of \textit{TaskType} to be selected - in Acme use case it can be \textit{SETaskTask} or \textit{CodingTaskType} or default value \textit{TaskType}).

The other tool, called \textit{Enacted process graphical editor}, provides process edition and tasks assignations through a graphical visualization displaying process being executed. This tool is represented on the right side of the figure \ref{fig:ToolingArchitecture}. It follows the same architecture pattern as explicited for the \textit{ProcessType graphical editor}, with two model slots referencing both the model and a diagram. Process enactment is operated from a \textit{ProcessType}, and must be defined using a name identifying the newly instantiated process. All tasks are created from their \textit{TaskType} definition, and required assignments applies (if for example "Coding" \textit{TaskType} defines COBOL as output programming language, related task is automatically assigned to "Ann Smith"). All tasks get a status as well as a set of performing actors, a begin and end date, and used and produced artifacts.

### 2021-05-27 8-Discussion.tex

  As for cross-level relationships, cross-level constraints do not have a
  specific nature. They are like any constraint a user can define on a model
  element. Therefore, if a constraint is mandatory when establishing a
  relationship between elements from different levels, the user has to create
  it manually. This can be a limit of our approach: the flexibility of our
  tooling at the cost of a limited number of automated behaviors.

@@

  In order to preserve the integrity of level contents when changes to level
  contents occur, a validation has to be triggered. The tool reports all
  inconsistencies that the user is free to fix. Note that in our approach, the
  metamodels are built in a ad hoc way together with the models
  (co-construction), therefore the user validates on a ad hoc basis.

@@

  \noteJC{pas trop clair pour moi pour le moment -> N/A}

@@

Our tooling includes mechanisms to verify the models. First, syntactical consistency is realised by cardinality checks and by typing. Second, the constraints the users have written are continuously checked in order to

However, a user can add verification mechanisms by writing new behaviors
attached to concepts. Presently, the verifications rely on the users when they
build the metamodels and the models.

\todo[inline]{verification de cohérence syntaxique (cardinalités, type des instances) et des contraintes métier (comportements écrits par les humains) à la volée + validation manuelle ; typage ? ; pas d'analyseur statique ; vérif de la structure des contraintes/bonne forme des contraintes par rapport à la structure des modèles (exemple binding pas valide)}

>>>>

  As for cross-level relationships, cross-level constraints do not have a
  specific nature. They are like any constraint a user can define by adding a behavior to a model element. Therefore, if a constraint is mandatory when establishing a
  relationship between elements from different levels, the user has to create
  it manually. This can be a limit of our approach: the cost of the flexibility of our tooling is a limited number of automated behaviors.

  %the flexibility of our tooling at the cost of a limited number of automated behaviors.

@@

  Behaviors are continuously checked, ensuring the integrity of the models when changes occur. 
  However, most of those behaviors have to be written by the users. 
  Thus, the integrity of the contents relies on the users when they build the metamodels and the models.
  Note that in our approach, the metamodels are built in a ad hoc way together with the models
  (co-construction), therefore the user also validates on a ad hoc basis.

@@

  \noteJC{pas trop clair pour moi pour le moment -> N/A ? }
  Due to the nature of our approach and its level-blindness, \emph{deep characterization} does not apply to our solution. \todo[inline]{je suppose qu'en fait, on pourrait encoder dans un comportement quelque chose se rapprochant}

@@

Our tooling includes mechanisms to verify the models. First, syntactical consistency is realised by cardinality checks and by typing. Second, the constraints the users have written are continuously checked. Thus, a part of the verification relies on the users when they build the metamodels and the models.
